---
title: "The *audiometry* package"
author: "Bernhard Lehnert, University of Greifswald, Germany"
date: "May 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro_to_PTA_Standards}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### In Short

If you deal with something like this, then this package is for you:


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, fig.dim=c(5,5)}
exmpl <- data.frame(f = c(125,250,500,1000,2000,3000,4000,8000),
                    t = c(15,10,25,35,40,55,75,60))
library(audiometry)
library(ggplot2)
gg_pta(data.frame(exmpl)) + 
  geom_point(aes(x=f, y=t), color="red", size=2.5) +
  geom_line(aes(x=f, y=t), color="red", lwd=.8)
```


### A bit longer

More then 30,000,000 people in the USA suffer from hearing loss, mostly the elderly, >80%  at  85  years  old  or  above. For more background and details on epidemiology see http://dx.doi.org/10.4236/ijohns.2013.24025 . Treatment is of utmost importance and great progress has been made in understanding, detecting as well as treating different conditions that lead to hearing loss, yet the situation is often dissatisfying and just a glance at the numbers of the affected, the huge impact on their lives and cost of some treatment options like cochlear implants serve as proof, that the topic deserves further research. 

Among the many approaches to measure hearing loss, pure tone audiometry (PTA) clearly plays the most important role, up to the point, where laymen call pure tone audiometry just "hearing test". 

In the simplest form, a patient is given headphones, sinus tones in different frequencies are played at different intensities and hearing thresholds are marked in a standardized diagram. Results are usually communicated visually via this diagram and thus it is only natural, to use the same diagram standards in scientific publication of pure tone audiometric data. 

### Standards to be met

There is an ISO standard ISO 8253-1:2010(en) that is not open, so the author could not refer to it. The American Speech-Languge-Hearing Association (ASHA) has published their guidelines openly ( https://www.asha.org/policy/GL2005-00014/#sec1.8 ) where they state:

> When the graphic form is used, the test frequencies shall be recorded on the abscissa, indicating frequency on a logarithmic scale, and hearing levels shall be recorded on the ordinate, using a linear scale to include the units of decibels. The aspect ratio of the audiogram is important for standardization. The correct aspect ratio is realized when a square is formed between any given octave pair on the abscissa and any 20 dB increment on the ordinate. For conventional audiometry, the vertical scale is to be designated hearing level in decibels; the horizontal scale is to be labeled frequency in hertz. By convention, frequency is recorded in ascending order from left to right, and hearing level is recorded in ascending order from top to bottom, ranging from a minimum value of −10 dB to the maximum output limits of the audiometer (usually 110 or 120 dB HL).

Whilst regarding all these rules is obviously possible in *ggplot2* it is not necessarily easy nor will every scientist in the field of hearing research always have these rules in mind. 

**That is, why the *audiometry* package provides an easy way to prepare a suitable *ggplot* object to start building diagrams on.**

### How to start

First, we define some example data with no particular meaning except for showing the functionality.

```{r}
ex1 <- data.frame(time = gl(3, 6),
                  f = rep(c(250, 500, 1000, 2000, 4000, 8000), 3),
                  t = c(60, 65, 40, 25, 10, 5, 
                        15, 25, 20, 10, 5, 0,
                        5, 5, 10, 0, 0, 10))
```


When starting a *ggplot2* graphic, you usually start by calling *ggplot* with a data.frame as in 

```{r}
library(ggplot2)
ggpl <- ggplot(ex1)
```

Instead we will start a pure-tone-audiogram-like plot with

```{r}
library(ggplot2)
library(audiometry)
ggpl <- gg_pta(ex1)
```

This will prepare our form. In the next step we need to add *geom*s to put our data into the form. Now this is all done with standard *ggplot2* commands. In our case, let's assume we optained a 6 frequency pure tone audiogram of a patient with sudden sensoryneural hearing loss in the deep frequencies at three different points in time: onset, two weeks and four weeks later, when she had totally recovered. We call *gg_pta* as our basis, then we add points with *geom_point* and lines with *geom_line* where line type and color represent the three points in time:

```{r, fig.dim=c(6,5)}
gg_pta(ex1) +
  geom_point(aes(x = f, y = t, color = time), size = 3, alpha=.5) +
  geom_line(aes(x = f, y = t, color = time, lty = time), lwd=1)
```

### Making changes

There are two ways to make changes to this plot. Either through *gg_pta* itself or employing all the manyfold possbilities *ggplot2* offers the adept.
If you wanted to change the language of the axis labels and the axis limits and the extra line at 0 dB to disapear, that can easily be done when calling *gg_pta*:

```{r, fig.dim=c(5,5)}
gg_pta(data.frame(0), xlab = "frequenza [Hz]", ylab = "volume [dB]",
       xlim = c(125, 12000), x_base_lwd = -1)
```

All the possible arguments to *gg_pta* are in the manual, just type
**help(gg_pta)**

### Marking sides

There is a number of ways to mark a side in PTA. *gg_pta* has a shortcut by means of the *lettermark* argument, which adds an "R" or "L" or both in size *lettermarksize* to the audiogram

```{r, fig.dim=c(5,5)}
gg_pta(data.frame(0), lettermark = c("R", "L"), lettermarksize = 40)
```


### Examples


Obviously, for the *ggplot2* savy, everything can be changed. Here are some examples with a randomly generated data set.
```{r, echo=FALSE}
id = gl(25, 9)
f = rep(c(250, 500, 750, 1000, 1500, 2000, 3000, 4000, 8000), 25)
t = c(replicate(13, c(20, 20, 20, 25, 25, 35, 40, 45, 40) +
                   sample(seq(-15,10,2.5),9, replace=TRUE)),
      replicate(12, c(10, 5, 10, 10, 10, 10, 15, 5, 15) +
                   sample(seq(-10,10,2.5),9, replace=TRUE)))
treat = c(rep("placebo", 15*9), rep("verum", 10*9))

some.data <- data.frame(id = id, freq = f, thres = t, treatment = treat)
```

```{r, fig.dim=c(5,5)}
gg_pta(some.data) + 
  geom_line(aes(x = freq, y = thres, group = id, color = treatment)) + 
  theme_grey() + 
  theme(legend.position = c(0.5, 0.2))
```



```{r, fig.dim=c(6,5)}
library(ggbeeswarm)
gg_pta(some.data, xlim=c(125, 10000), x_base_lwd = 1.5,
       xlab = "Frequenz [Hz]", ylab = "Lautstärkepegel [dB HL]") +
  geom_beeswarm(aes(x = freq, y = thres, color=treatment), cex = 1.5) +
  theme_linedraw()
```

In order to decide, whether insurances will pay for hearing aids, German ENT specialists have to check, whether at least two thresholds in between 0.5 kHz an 4 kHz are 30 dB or louder. That decision line is easily visualized:

```{r, fig.dim=c(5,5)}
library(ggthemes)
gg_pta(data.frame(0)) +
  geom_segment(aes(x = 500, y = 29, xend = 4000, yend = 29), colour = "red", lwd=2) +
  geom_segment(aes(x = 500, y = 31, xend = 4000, yend = 31), colour = "green", lwd=2) +
  theme_gdocs()

```

### Finally

The package name *audiometry* is obviously much wider then just pure tone audiogram forms.  I intend to broaden the package's functionality *slowly* over time. As of now the *audiometry* packages has support for the Freiburger Sprachtest which is a widely used speech audiometry test in German speaking countries.



-- Bernhard Lehnert
